### 2014-09-30 paper.tex
updated some figure descriptions

% @CTB why are the global ones not included in the final gene models?
% @LP they are included in the final models if they align to the genome.

@@

% @CTB how many reads are discarded in the process of mapping?

@@

% @LP Maybe I should add bitscores from Cufflinks so that we can compare with
% those from Gimme. If the results are similar, we can say that extending
% Cufflinks with assembly does not affect gene structures in a bad way.
To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
50 or more amino acids.  We then searched for homologous sequences in mouse
ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
genes match mouse proteins at a bit-score/length $\ge1.0$
(Fig.~\ref{bitscore}).  A bit-score/length ratio greater than 1 indicates that
approximately more than half of the protein sequence matches mouse proteins
well. The results suggest that merging gene models from Cufflinks and {\em de
novo} aseembly did not disrupt the structure of protein sequences.

% @CTB for discussion: what's surprising about all of this? so many
% splice variants, etc?  Note we need to go over pickrell paper carefully.
% @LP I discuss about it below. Hope it helps.

@@

\subsection*{Mapping reads to The Genome and Gene Models}

@@

Figure~\ref{overall_pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv}. Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).

@@

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{overall_pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two
assembly methods -- global and local assembly.  Transcripts are aligned to the
chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
of transcripts.  }
\label{overall_pipeline}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{local_assembly}
\end{center}
\caption{
{\bf Local Assembly Pipeline.}
Reads are first mapped to a chicken genome.  Then only mapped reads are
assembled by Velvet and Oases.
}
\label{local_assembly}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{algorithm.pdf}
\end{center}
\caption{
{\bf Intron and exon graphs.} Each intron connects to exons whose splice
junctions match it boundary.  Some exons are excluded from the final gene model
if they are incomplete (exon 3a,b).  Introns sharing at least one exon are
grouped together.  Then an exon graph is made using exons as nodes.
}
\label{algorithm}
\end{figure}

@@

\end{document}

>>>>

@@

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
translated 28,772 of 34,800 (82.6\%) of isoforms from Gimme models to protein
sequences with 50 or more amino acids.  We then searched for homologous
sequences in mouse ENSEMBL, and found that 22,460 (77.7\%) and 22,991 (79.9\%)
of isoforms from Cufflinks and Gimme match mouse proteins at a bit-score/length
$\ge1.0$(Fig.~\ref{bitscore}).  The number of distinct genes are 16,731 and
12,945 from Cufflinks and Gimme respectively.  A bit-score/length ratio greater
than 1 indicates that approximately more than half of the protein sequence
matches mouse proteins well. The sizes of genes and isoforms from Gimme are
slightly greater than those from Cufflinks while the bitscore/length ratios are
slightly lower suggesting that transcripts from assembly are more complete.

% @CTB for discussion: what's surprising about all of this? so many splice
% variants, etc?  Note we need to go over pickrell paper carefully.  @LP I
% discuss about it below. Hope it helps.

@@

\subsection*{Mapping Reads to The Genome and Gene Models}

@@

Figure~\ref{img:pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv}. Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).

@@

\section*{Figure Legends}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.}
Transcripts are obtained from two assembly methods -- global and
local assembly. Transcripts with low complexity are removed and
poly(A) tails are trimmed by SeqClean. Transcripts are aligned to
the chicken genome using BLAT and alignments are then used by
Gimme to construct gene models. Gene models from Cufflinks and
other sources can also be incorporated to build the gene models.
}
\label{img:pipeline}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{local_assembly.pdf}
\end{center}
\caption{
{\bf Local assembly pipeline.}
Reads are first mapped to a chicken genome by Tophat.  Then reads
mapped to each chromosome are assembled by Velvet and Oases
separately.
}
\label{local_assembly}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{algorithm.pdf}
\end{center}
\caption{
{\bf Intron and exon graph.}
Introns sharing at least one exon are grouped together, then an
exon graph is constructed with exons from all introns as nodes.
Exons are excluded from the final gene model if they are smaller
than other exons sharing the same intron (exon 1a, 3a and 3b).
Edges of the exon graph (red arrow) are implied from the
connections of introns (blue arrow).  A transcript is constructed
from concatenation of all unique exons in the graph.
}
\label{algorithm}
\end{figure}
\end{document}
